---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(timeout = 2000)
```

# EMODnetWFS: Access EMODnet Web Feature Service data through R

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R build status](https://github.com/EMODnet/EMODnetWFS/workflows/R-CMD-check/badge.svg)](https://github.com/EMODnet/EMODnetWFS/actions)
[![Codecov test coverage](https://codecov.io/gh/EMODnet/EMODnetWFS/branch/master/graph/badge.svg)](https://codecov.io/gh/EMODnet/EMODnetWFS?branch=master)
<!-- badges: end -->

The goal of EMODnetWFS is to allow interrogation of and access to EMODnet geographic vector data in R though the [EMODnet Web Feature Services](https://www.emodnet.eu/en/data). [Web Feature services (WFS)](https://www.ogc.org/standards/wfs) represent a change in the way geographic information is created, modified and exchanged on the Internet and offer direct fine-grained access to geographic information at the feature and feature property level. This package was developed by the Sheffield University during the EMODnet Biology WP4 data products workshop in June 2020.

You can read the product story on the EMODnet-Biology portal following [this link](https://www.emodnet-biology.eu/blog/emodnetwfs-access-emodnet-web-feature-service-data-through-r), or read the vignette directly in R after installing the package.

```{r, eval=FALSE}
vignette("emodnetwfs")
```


## Installation

You can install the development version of EMODnetWFS from GitHub with:

``` r
remotes::install_github("EMODnet/EMODnetWFS")
```

## Available services

All available services are contained in the tibble returned by `emodnet_wfs()`.


```{r, echo=FALSE}
library(EMODnetWFS)
knitr::kable(emodnet_wfs())
```

To explore available services in Rstudio use:

```r
View(emodnet_wfs())
```

**On Linux, the geology services will not work without tweaking your openssl setting (e.g. see the second suggestion in https://askubuntu.com/questions/1233186/ubuntu-20-04-how-to-set-lower-ssl-security-level/1233456#1233456). This approach is for expert users only, be cautious. See https://github.com/EMODnet/EMODnetWFS/issues/69 for more details and updates.**

## Create Service Client

Create new WFS Client.  Specify the service using the `service` argument. 

```{r}
wfs_bio <- emodnet_init_wfs_client(service = "biology")

wfs_bio$getUrl()
```

## Get WFS Layer info

You can get metadata about the layers available from a service. 

```{r}
emodnet_get_wfs_info(service = "biology")
```

or you can pass a wfs client object.

```{r}
emodnet_get_wfs_info(wfs_bio)
```

You can also get info for specific layers from wfs object:

```{r}
wfs_cml <- emodnet_init_wfs_client("biology")
emodnet_get_wfs_info(wfs_cml)

layers <- c("mediseh_zostera_m_pnt","mediseh_posidonia_nodata")

emodnet_get_layer_info(wfs = wfs_cml, layers = layers)
```


Finally, you can get details on all available services and layers from the server

```{r eval=FALSE}
emodnet_get_all_wfs_info()
```


## Get WFS layers

You can extract layers directly from a `wfs` object using layer names. All layers are downloaded as `sf` objects and output as a list with a named element for each layer requested.

```{r}
emodnet_get_layers(wfs = wfs_cml, layers = layers)
```

You can change the output `crs` through argument `crs`.

```{r}
emodnet_get_layers(wfs = wfs_cml, layers = layers, crs = 3857)

```

You can also extract layers directly from a WFS service.

```{r}
emodnet_get_layers(
    service = "biology",
    layers = c("mediseh_zostera_m_pnt", "mediseh_posidonia_nodata")
)

```


Layers can also be returned to a single `sf` object through argument `reduce_layers`.  If `TRUE` the function will try to reduce all layers into a single `sf`.

If attempting to reduce fails, it will return a list with a warning:
 
```{r, error=TRUE}
emodnet_get_layers(
    wfs = wfs_cml, 
    layers = layers,
    reduce_layers = TRUE
)
```

Using `reduce_layers = TRUE` is also useful for returning an `sf` object rather than a list in single layer request.

```{r}
emodnet_get_layers(
    service = "biology",
    layers = c("mediseh_posidonia_nodata"), 
    reduce_layers = TRUE
)
```


## Citation

Please cite this package as:

Anna Krystalli (2020). EMODnetWFS: Access EMODnet Web Feature Service data through R. R package version 0.0.2. https://github.com/EMODnet/EMODnetWFS. Integrated data products created under the European Marine Observation Data Network (EMODnet) Biology project (EASME/EMFF/2017/1.3.1.2/02/SI2.789013), funded by the by the European Union under Regulation (EU) No 508/2014 of the European Parliament and of the Council of 15 May 2014 on the European Maritime and Fisheries Fund.
