---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EMODnetWFS

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of EMODnetWFS is to allow interrogation and access to the EMODnet Web Feature Services data in R.

## Installation

You can install the released version of EMODnetWFS from [CRAN](https://CRAN.R-project.org) with:

``` r
remotes::install_github("annakrystalli/EMODnetWFS")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(EMODnetWFS)
## basic example code
```

Create new WFS Client. The default service is `https://ows.emodnet-seabedhabitats.eu/emodnet_open_maplibrary/wfs`

```{r}
wfs <- emodnet_init_wfs_client()

```

You can access further services through the `service`. 

```{r}

wfs_bath <- emodnet_init_wfs_client(service = "https://ows.emodnet-bathymetry.eu/wfs")

wfs_bath$getUrl()
```

## Get WFS Layer info

```{r}
emodnet_get_wfs_info()
```

You can access information about a service using the `service` argument.

```{r}
emodnet_get_wfs_info(service = "https://ows.emodnet-bathymetry.eu/wfs")
```

or you can pass a wfs client object.

```{r}
emodnet_get_wfs_info(wfs_bath)
```

## Get WFS layers

```{r}
emodnet_get_layers(layers = c("dk003069", "dk003070"))

```

Layers can also be returned to a single `sf` object through argument `reduce_layers`. If `TRUE` the function will try to reduce all layers into a single `sf` if possible.
```{r}

emodnet_get_layers(layers = c("dk003069", "dk003070"), reduce_layers = TRUE)
```

```{r}
#emodnet_get_layers(wfs_bath, layers = "source_references_2016")
```

