---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EMODnetWFS <img src='man/figures/emodnetwfs.svg' align="right" height="139"/>

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R build status](https://github.com/EMODnet/EMODnetWFS/workflows/R-CMD-check/badge.svg)](https://github.com/EMODnet/EMODnetWFS/actions)
[![Codecov test coverage](https://codecov.io/gh/EMODnet/EMODnetWFS/branch/master/graph/badge.svg)](https://codecov.io/gh/EMODnet/EMODnetWFS?branch=master)
<!-- badges: end -->

The goal of EMODnetWFS is to allow interrogation and access to the EMODnet Web Feature Services data in R.

## Installation

You can install the development version of EMODnetWFS from GitHub with:

``` r
remotes::install_github("EMODnet/EMODnetWFS")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(EMODnetWFS)
## basic example code
```

### Available services

All available services are contained in the `emodnet_wfs` package dataset.


```{r, echo=FALSE}
knitr::kable(emodnet_wfs)
```

To explore available services in Rstudio use:
```r
View(emodnet_wfs)
```

### Create Service Client

Create new WFS Client. The default service is `seabed_habitats_individual_habitat_map_and_model_datasets`.

```{r}
wfs <- emodnet_init_wfs_client()

```

You can access further services using the `service` argument. 

```{r}

wfs_bath <- emodnet_init_wfs_client(service = "bathymetry")

wfs_bath$getUrl()
```

## Get WFS Layer info

You can get metadata about the layers available from a service. The default service is `seabed_habitats_individual_habitat_map_and_model_datasets`.

```{r}
emodnet_get_wfs_info()
```

You can access information about a service using the `service` argument.

```{r}
emodnet_get_wfs_info(service = "bathymetry")
```

or you can pass a wfs client object.

```{r}
emodnet_get_wfs_info(wfs_bath)
```

You can also get info for specific layers from wfs object:

```{r}
wfs_cml <- emodnet_init_wfs_client("chemistry_marine_litter")
emodnet_get_wfs_info(wfs_cml)

layers <- c("bl_fishing_cleaning",
          "bl_beacheslocations_2001_2008_monitoring")

emodnet_get_layer_info(wfs = wfs_cml, layers = layers)
```


Finally, you can get details on all available services and layers from the server

```{r eval=FALSE}
emodnet_get_all_wfs_info()
```


## Get WFS layers

You can extract layers directly from a `wfs` object using layer names. All layers are downloaded as `sf` objects and the crs of outputs are standardised to `EPSG` code 4326 by default.

```{r}
emodnet_get_layers(wfs = wfs_cml, layers = layers)
```

You can chage the output crs through argument `crs`.

```{r}
emodnet_get_layers(wfs = wfs_cml, layers = layers, crs = 3857)

```


You can also extract layers directly from a WFS service The default service is `seabed_habitats_individual_habitat_map_and_model_datasets`.

```{r}
emodnet_get_layers(layers = c("dk003069", "dk003070"))

```

Use argument `service` to specify the required service.
```{r}

human_activities <- emodnet_get_layers(service = "human_activities", 
                   layers = c("aquaculture", "dredging"))

human_activities[["aquaculture"]]
```


Layers can also be returned to a single `sf` object through argument `reduce_layers`.  If `TRUE` the function will try to reduce all layers into a single `sf` and will fail if not.

```{r}
emodnet_get_layers(layers = c("dk003069", "dk003070"), 
                   reduce_layers = TRUE)
```

 
```{r, error=TRUE}
emodnet_get_layers(wfs = wfs_cml, layers = layers,
                   reduce_layers = TRUE)
```


