---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(timeout = 2000)
```

# EMODnetWFS: Access EMODnet Web Feature Service data through R

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![R build status](https://github.com/EMODnet/EMODnetWFS/workflows/R-CMD-check/badge.svg)](https://github.com/EMODnet/EMODnetWFS/actions)
[![Codecov test coverage](https://codecov.io/gh/EMODnet/EMODnetWFS/branch/main/graph/badge.svg)](https://app.codecov.io/gh/EMODnet/EMODnetWFS/tree/main)
<!-- badges: end -->

The goal of EMODnetWFS is to allow interrogation of and access to [EMODnet geographic vector data](https://emodnet.ec.europa.eu/en/emodnet-web-service-documentation#inline-nav-3) in R though the [EMODnet Web Feature Services](https://emodnet.ec.europa.eu/en/data). [Web Feature services (WFS)](https://www.ogc.org/standard/wfs/) represent a change in the way geographic information is created, modified and exchanged on the Internet and offer direct fine-grained access to geographic information at the feature and feature property level. 
EMODnetWFS aims at offering an user-friendly interface to this rich data.

## Installation

You can install the development version of EMODnetWFS from GitHub with:

``` r
# install.packages("remotes")
remotes::install_github("EMODnet/EMODnetWFS")
```

## Available services

All available services are contained in the tibble returned by `emodnet_wfs()`.


```{r, echo=FALSE}
library(EMODnetWFS)
knitr::kable(emodnet_wfs())
```

To explore available services in Rstudio use:

```r
View(emodnet_wfs())
```

## Create Service Client

Create new WFS Client.  Specify the service using the `service` argument. 

```{r}
wfs_bio <- emodnet_init_wfs_client(service = "biology")

wfs_bio
```

## Get WFS Layer info

You can get metadata about the layers available from a service. 

```{r}
emodnet_get_wfs_info(service = "biology")
```

or you can pass a wfs client object.

```{r}
emodnet_get_wfs_info(wfs_bio)
```

You can also get info for specific layers from wfs object:

```{r}
layers <- c("mediseh_zostera_m_pnt", "mediseh_posidonia_nodata")

emodnet_get_layer_info(wfs = wfs_bio, layers = layers)
```


Finally, you can get details on all available services and layers from the server

```{r eval=FALSE}
emodnet_get_all_wfs_info()
```


## Get WFS layers

You can extract layers directly from a `wfs` object using layer names. 
All layers are downloaded as `sf` objects and output as a list with a named element for each layer requested.

```{r}
emodnet_get_layers(wfs = wfs_bio, layers = layers)
```

You can change the output `crs` through the argument `crs`.

```{r}
emodnet_get_layers(wfs = wfs_bio, layers = layers, crs = 3857)
```

You can also extract layers using a WFS service name.

```{r}
emodnet_get_layers(
  service = "biology",
  layers = c("mediseh_zostera_m_pnt", "mediseh_posidonia_nodata")
)
```


Layers can also be returned to a single `sf` object through argument `reduce_layers`.  
If `TRUE` the function will try to reduce all layers into a single `sf`.

If attempting to reduce fails, it will error:
 
```{r, error=TRUE}
emodnet_get_layers(
  wfs = wfs_bio,
  layers = layers,
  reduce_layers = TRUE
)
```

Using `reduce_layers = TRUE` is also useful for returning an `sf` object rather than a list in single layer request.

```{r}
emodnet_get_layers(
  service = "biology",
  layers = c("mediseh_posidonia_nodata"),
  reduce_layers = TRUE
)
```

## Citation

To cite EMODnetWFS, please use the output from `citation(package = "EMODnetWFS")`.

```{r}
citation(package = "EMODnetWFS")
```

## Acknowledgements

This package was started by the Sheffield University during the EMODnet Biology WP4 data products workshop in June 2020.
You can read the [product story on the EMODnet-Biology portal](https://emodnet.ec.europa.eu/geonetwork/)
