---
title: "EMODnetWFS Case Study: Accessing and mapping EMODnet data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EMODnetWFS Case Study: Accessing and mapping EMODnet data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, collapse = TRUE, comment = "#>")
```

```{r, include=FALSE}
library(httptest)
start_vignette("emodnetwfs-case-study")
```

## Introduction

The package was designed to make EMODnet vector data layers easily accessible in R. The package allows users to query information on and download data from all available [EMODnet Web Feature Service (WFS) endpoints](https://www.emodnet.eu/en/data) directly into their R working environment. Data are managed as [`sf` objects](https://r-spatial.github.io/sf/) which are currently the state-of-the-art in handling of vector spatial data in R. The package also allows user to specify the coordinate reference system of imported data.

## Data Product
### Installation

You can install the development version of `EMODnetWFS` from GitHub with:

```{r install, eval = FALSE}
remotes::install_github("EMODnet/EMODnetWFS")
```

### Explore the EMODnet WFS services with R

For this tutorial we will make use of the `sf`,  `dplyr`  and `mapview` packages. The simple features `sf` package is a well known standard for dealing with geospatial vector data. The package `dplyr` is a strong library for data manipulation. This package also loads `magickr`'s pipe operator `%>%`, which allows to write pipelines in R. To visualize geometries,  `mapview` will create quick interactive maps.

Run this line to install these packages:

```{r install.packages, eval = FALSE}
install.packages(c("sf", "dplyr", "mapview"))
```



With the `EMODnetWFS` package, we can explore and combine the data served by the [EMODnet lots ](https://www.emodnet.eu/en/portals) through [OGC Web Feature Services](https://en.wikipedia.org/wiki/Web_Feature_Service) or WFS. 

Imagine we are interested on seabed substrates. The first stop is to choose what EMODnet lot can provide with these data. For that, we can check the services available on the `emodnet_wfs` dataset contained inside the package. 

```{r error = FALSE, warning = FALSE, message = FALSE}
library(EMODnetWFS)
library(mapview)
library(dplyr)
library(sf)

emodnet_wfs
```



The column `service_name` shows services available, while  `service_url` has the corresponding base url to perform a WFS request. The Seabed portal should have the data we are looking for. A WFS client can be created by passing the corresponding `service_name` to the function `emodnet_init_wfs_client()`. The layers available to this WFS client are consulted with `emodnet_get_wfs_info()`.

```{r}
seabed_wfs_client <- emodnet_init_wfs_client(service = "seabed_habitats_general_datasets_and_products")

emodnet_get_wfs_info(wfs = seabed_wfs_client)
```



Each layer is explained in the `abstract` column. We can see several layers with the information provided by the EU member states for the [Habitats Directive 92/43/EEC reporting](https://www.eea.europa.eu/data-and-maps/data/article-17-database-habitats-directive-92-43-eec-2). We will select the layers about coastal lagoons, mudflats and sandbanks with their respective `layer_name`.

```{r}
habitats_directive_layer_names <- c("art17_hab_1110", "art17_hab_1140", "art17_hab_1150")

emodnet_get_layer_info(wfs = seabed_wfs_client, layers = habitats_directive_layer_names)
```



We are now ready to read the layers into R with `emodnet_get_layers()`. EMODnetWFS reads the geometries as simple features (See `sf` package) transformed to [4326](https://epsg.io/4326) by default. Specifying another map projection is possible by passing a EPGS code or projection string with `emodnet_get_layers(crs = "your projection")`. The argument `reduce_layers = TRUE` stack all the layers in one single tibble. Default is FALSE and returns a list of sf objects, one per layer.

```{r}
habitats_directive_layers <- emodnet_get_layers(wfs = seabed_wfs_client, 
                                                layers = habitats_directive_layer_names, 
                                                reduce_layers = TRUE)  

class(habitats_directive_layers)

glimpse(habitats_directive_layers)
```



Run the following code to have a quick look at the layers geometries
```{r, eval = TRUE}
# Transform to Polygon geometry type from Multisurface
if(unique(st_geometry_type(habitats_directive_layers)) == "MULTISURFACE"){
  habitats_directive_layers <- habitats_directive_layers %>% 
    st_cast(to = "GEOMETRYCOLLECTION") %>% 
    st_collection_extract(type = "POLYGON")
}

# Visualize
map <- mapview(habitats_directive_layers, zcol = "habitat_description", burst = TRUE)

map
```


Furthermore, we can get data from other EMODnet lots and combine them. The Human Activities portal provides the maritime boundaries of the European Union state members. This time we will not initiate a WFS client, but we will call the service directly. The WFS client will be generated on the fly.

Same as before, we have a look at the layers available first.

```{r}
emodnet_get_wfs_info(service = "human_activities")
```



The `layer_name` for the [maritime boundaries](https://www.eea.europa.eu/data-and-maps/data/maritime-boundaries) seems to be `maritimebnds`. This dataset was developed based on the official [data provided by the European Environmental Agency](https://www.eea.europa.eu/data-and-maps/data/maritime-boundaries) and the [Maritime Boundaries Database](https://marineregions.org/eez.php) compiled by MarineRegions.org (Flanders Marine Institute, 2019).

The `sitename` variable specifies the type of boundary each feature represents. For illustration purposes, we will filter our request to return only [Territorial Seas](https://www.un.org/depts/los/convention_agreements/texts/unclos/part2.htm). 

```{r}
maritime_boundaries <- emodnet_get_layers(service = "human_activities", 
                                          layers = "maritimebnds", 
                                          reduce_layers = TRUE,
                                          cql_filter = "sitename='Territory sea (12 nm)'")

glimpse(maritime_boundaries)
```



Add the maritime boundaries to the previous map with this line. To do so, cats to `MULTILINESTRING` geometry type, transform to the same crs as the habitats directive `sf` (only required because of the next step), and crop using the bounding box of the habitats directive `sf`.
```{r, eval = TRUE}

maritime_boundaries <- maritime_boundaries %>% 
  st_cast(to = "MULTILINESTRING") %>%
  st_transform(st_crs(habitats_directive_layers)) %>%
  st_crop(st_bbox(habitats_directive_layers))

map + mapview(maritime_boundaries )
```


We have now combined data from the Seabed Habitats and Human Activities portals. However, there is more! EMODnet provides also physics, chemistry, biological or bathymetry data. Explore all the layers available with.

```{r, eval = FALSE}
emodnet_get_all_wfs_info()
```



## More information

### References

Blondel, Emmanuel. (2020, May 27). ows4R: R Interface to OGC Web-Services (Version 0.1-5). Zenodo. http://doi.org/10.5281/zenodo.3860330

Flanders Marine Institute (2019). Maritime Boundaries Geodatabase, version 11. Available online at https://www.marineregions.org/. https://doi.org/10.14284/382.

Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2020). dplyr: A Grammar of Data Manipulation. R package version 1.0.2.https://CRAN.R-project.org/package=dplyr

Pebesma E (2018). “Simple Features for R: Standardized Support for Spatial Vector Data.” The R Journal, 10(1), 439–446. doi: 10.32614/RJ-2018-009, https://doi.org/10.32614/RJ-2018-009. 

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

Tim Appelhans, Florian Detsch, Christoph Reudenbach and Stefan Woellauer (2020). mapview: Interactive Viewing of Spatial Data in R. R package version 2.9.0. https://CRAN.R-project.org/package=mapview

### Code

Please cite this package as: 

Anna Krystalli (2020). EMODnetWFS: Access EMODnet Web Feature Service data through R. R package version 0.0.2. https://github.com/EMODnet/EMODnetWFS. Integrated data products created under the European Marine Observation Data Network (EMODnet) Biology project (EASME/EMFF/2017/1.3.1.2/02/SI2.789013), funded by the by the European Union under Regulation (EU) No 508/2014 of the European Parliament and of the Council of 15 May 2014 on the European Maritime and Fisheries Fund.

```{r, include=FALSE}
end_vignette()
```

