---
title: "Using request parameters to limit query results"
---


Requesting a full layer with `emodnet_get_layers()` will return the geometries for all features in that layer, potentially a large amount of data. 

Some basic ability to limit the output of any query is possible by providing additional parameters to the feature request query. This can be achieved by passing additional arguments to `emodnet_get_layers(...)`.

For details on all available parameters consult the [`GetFeature` Geoserver documentation](https://docs.geoserver.org/master/en/user/services/wfs/reference.html#getfeature).  Also note that there are [differences in available parameters between different version of WFS specifications](https://docs.geoserver.org/stable/en/user/services/wfs/basics.html#differences-between-wfs-versions). We recommend and mainly document using the default and latest version (`"2.0.0"`) which is supported by all EMODnet WFS services.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Initialise a WFS client

Let's start by loading emodnet.wfs and initialising a WFS client to the `human_activities` WFS service.

```{r setup}
library(emodnet.wfs)

wfs <- emodnet_init_wfs_client("human_activities")
```


## Return single feature using feature ID

We can restrict the query to a single feature using `featureID` and providing the ID of a specific feature.

```{r}
emodnet_get_layers(
  wfs = wfs,
  layers = "pipelines",
  featureID = "pipelines.1",
  reduce_layers = TRUE
)
```

## Return specific number of features 

If the ID of the feature is unknown but we still want to limit the number of features returned, we use the `count` parameter for WFS 2.0.0 or the `maxFeatures` parameter for earlier WFS versions to restrict the number of features returned. In this example we restrict the query to the first feature.

```{r}
emodnet_get_layers(
  wfs = wfs,
  layers = "pipelines",
  count = 1L,
  reduce_layers = TRUE
)
```

## Order features

Exactly which features will be returned depends in the internal structure of the data. However, we can sort the returned selection based on an attribute value using the `sortBy` parameter. 

Say we wanted to return 5 of the longest marine pipelines. In the following example, we sort by the `"length_m"` attribute and combine it with the `count` parameter to restrict the number of features returned. This returns the first 5 features after the data has been ordered by `"length_m"`. Some pipelines do not have their lengths recorded, we will skip them with a CQL Filter.

```{r}
emodnet_get_layers(
  wfs = wfs,
  layers = "pipelines",
  sortBy = "length_m+D",
  count = 5L,
  reduce_layers = TRUE,
  cql_filter = "length_m IS NOT NULL"
)
```


The default sort operation is to sort in ascending order. Some WFS servers require the sort order to be specified. In this case, append a `+A` to the attribute character string passed to `sortBy`. Conversely, here we add a `+D` to sort in descending order.

## Return blocks of features from specific starting point

Finally in WFS version 2.0.0 (and also available in earlier versions on GeoServer) a `startIndex` parameter was introduced, allowing users to specify the starting index of features to be returned.

```{r, echo=FALSE}
n_features <- layer_attributes_tbl(wfs = wfs, layer = "pipelines") %>% nrow()

n_features
```

For example, the full `"pipelines"` layer contains `r n_features` features. 

```{r, eval=FALSE}
n_features <- layer_attributes_tbl(wfs = wfs, layer = "pipelines") %>% nrow()

n_features
```


Setting the `startIndex` to `n_features - 4` (`r n_features - 4`) returns the last 4 features.

```{r}
startIndex <- n_features - 4L
startIndex

emodnet_get_layers(
  wfs = wfs,
  layers = "pipelines",
  startIndex = startIndex,
  reduce_layers = TRUE
)
```


We can also combine `startIndex` and `count` to return specific blocks of features. For example, in the following query we request features 5-9 by supplying a `startIndex` of 4 and  `count` of 5.

Note that `startIndex` uses `0` as a starting index (0 indicates the first feature), hence we set it to `4` in order to start at the 5th feature.

```{r}
emodnet_get_layers(
  wfs = wfs,
  layers = "pipelines",
  startIndex = 4L,
  count = 6L,
  reduce_layers = TRUE
)
```

This in combination with, for example, a `for` loop, can be used to download and process large layer which might not be able to be handled in R in smaller, more manageable blocks of features.


## Return data for specific attributes

Finally, we can specify specific attributes for which we want data returned using parameter `propertyName`.

To limit data returned to the single attribute `"country"` from just 3 features, we use parameters `propertyName = "country"` and `count = 3`. 

Note that when limiting data to specific attributes, all other columns are returned as `NA` (including any spatial, in this case `the_geom`, column).

```{r}
emodnet_get_layers(
  wfs = wfs,
  layers = "pipelines",
  propertyName = "country",
  count = 3L,
  reduce_layers = TRUE
)
```

To limit to multiple attributes, separate each attribute name in the character string supplied to `propertyName` by a comma.

```{r}
emodnet_get_layers(
  wfs = wfs,
  layers = "pipelines",
  propertyName = "country,the_geom",
  count = 3L,
  reduce_layers = TRUE
)
```


## Limit spatial extent using a boundary box

The `bbox` parameter allows us to search for features that are contained (or partially contained) inside a box of user-defined coordinates. The format of the `bbox`  parameter in most EMODnet WFS cases (assuming you are using version `2.0.0`) is `bbox=a1,b1,a2,b2,[crs]` where `a1`, `b1`, `a2`, and `b2` represent the coordinate values. The optional `crs` parameter is used to name the coordinate reference system (CRS) for the bbox coordinates (if they are different to the featureTypes native CRS.) The order of coordinates passed to the `bbox` parameter depends on the coordinate system used (this is why the coordinate syntax isn’t represented with `x` or `y`).

In the following example, we specify the bounding box by proving coordinates in the order `xmin,ymin,xmax,ymax` and specifying that coordinates are given in the `EPSG:4326` coordinate reference system.

```{r}
bbox_response <- emodnet_get_layers(
  service = "biology",
  layers = "mediseh_posidonia_nodata",
  bbox = "22.9,34.4,26.8,35.8,EPSG:4326",
  reduce_layers = TRUE
)

bbox_response %>%
  sf::st_cast(to = "MULTILINESTRING") %>%
  mapview::mapview(burst = TRUE, legend = FALSE)
```


## Non-standard Vendor Parameters

WFS vendor parameters are additional, non-standard request parameters defined by an implementation to provide enhanced capabilities. 

The majority of EMODnet services are GeoServer WFS implementations which support a variety of vendor-specific WFS parameters.

One of these are **`cql_filters`**. These are handled explicitly through the emodnet.wfs package through the `cql_filter` argument in `emodnet_get_layers()` and are documented in detail in `article("ecql_filtering")`.

For additional vendor parameters available through GeoServer implementations please refer to the [WFS vendor parameters documentatio](https://docs.geoserver.org/latest/en/user/services/wfs/vendor.html)

Exceptions are the Chemistry (Eutrophication by sea region, Eutrophication) services which uses Python and Chemistry (Litter) which uses MapServer so vendor parameters for these services might differ.


